\name: A flexible and data-driven K-process model for nucleosynthesis

### 2023-10-16 Paper/main.tex

\documentclass[modern]{aastex631}
\usepackage{amsmath}
\usepackage{microtype}
\usepackage{soul}

@@

\begin{abstract}\noindent
The element abundance pattern found by APOGEE and GALAH in Milky Way disk stars is close to two-dimensional, dominated by production from one prompt process and one delayed process.
This simplicity is remarkable, since the elements are produced by a multitude of nucleosynthesis mechanisms operating in stars with a wide range of progenitor masses.
We fit the abundances of 14 elements for 48,659 red-giant stars from APOGEE DR17 using a flexible, data-driven $K$-process model---dubbed \name.
In our fiducial model, with $K=2$, each abundance in each star is described as the sum of a prompt and a delayed process contribution.
We find that \name{} with $K=2$ is able to explain the abundances well, recover the observed abundance bimodality, and detect the bimodality over a greater range in metallicity than previously has been possible.
We compare to prior work by \citet{weinberg2022}, finding that \name{} produces similar results, but that \name{} better predicts stellar abundances, especially for elements C+N and Mn and for stars at super-solar metallicities.
The model makes assumptions, including especially that it fixes some parameters to break degeneracies and improve interpretability; we find that some of the nucleosynthetic implications are dependent upon these detailed parameter choices.
We add a third and fourth process (to make $K=4$), finding that the additional processes give the model more freedom and improve the model's ability to predict the stellar abundances, as expected, but they don't qualitatively change the story. 
The results of \name{} have implications for the formation of the Galaxy disk, the relationship between abundances and ages, and the physics of nucleosynthesis.
\end{abstract}

@@

\begin{deluxetable}{cccc}
    \tablecaption{Fiducial model $\Acc$ and $\AIa$ values for our stellar sample.
    \label{tab:qIa}}
    \tablehead{
\colhead{APOGEE ID} &	\colhead{[Mg/H]} & \colhead{$\Acc$} &	\colhead{$\AIa$}} 
\startdata
2M00000546+6152107 &	-0.20 &	0.61 &	0.48\\
2M00000866+7122144 &	-0.10 &	0.82 &	0.69\\
2M00001328+5725563 &	0.04 &	1.09 &	1.02\\
2M00001653+5540107 &	0.06 &	1.13 &	0.36\\
2M00001717+6147500 &	-0.23 &	0.60 &	0.39\\
... & ... & ... & ...\\
\enddata
\tablecomments{Full table available online.}
\end{deluxetable}

@@

Such analysis with \name{} parameters will be useful in studying nucleosynthesis, dynamics, disk formation, stellar ages, and much more. However, in this paper we only present fits for a small population with restricted stellar parameters, relative to the full APOGEE sample. 
\ejg{Add references to tables with q values at knots and A values for all stars.} While \name{} could be fit to the full APOGEE sample, systematic abundance effects with $\teff$ and $\logg$, as well as other abundance artifacts \citep[e.g.,][]{jonsson2020, griffith2021a}, cause the abundance trends to differ across the Hertzsprung-Russell diagram. The best-fit \name{} parameters for the giants would differ from those for the dwarfs. If such systematics could be accounted for (see Sit et al. in prep) we could fit the full APOGEE stellar sample with \name{}, or train \name{} on a subset of high signal-to-noise stars and apply the fits to the full sample. This potential future analysis could reveal additional information about the nucleosynthetic history of our Galaxy and would provide higher signal-to-noise abundance labels for the full sample.

>>>>

\documentclass[modern, linenumbers]{aastex631}
\usepackage{amsmath}
\usepackage{microtype}
\usepackage{soul}
% \usepackage{lineno}
% \linenumbers

@@

\begin{abstract}\noindent
The element abundance pattern found by APOGEE in Milky Way disk stars is close to two-dimensional, dominated by production from one prompt process and one delayed process.
This is remarkable, since the elements are known to be produced by many more than two nucleosynthetic sources.
We fit the abundances of 14 elements for 48,659 red-giant stars from APOGEE DR17 using a flexible, data-driven $K$-process model---dubbed \name.
In our fiducial model, with $K=2$, each abundance in each star is described as the sum of a prompt and a delayed process contribution.
We find that \name{} with $K=2$ is able to explain the abundances well, recover the observed abundance bimodality, and detect the bimodality over a greater range in metallicity than previously has been possible.
We compare to prior work by \citet{weinberg2022}, finding that \name{} produces similar results, but that \name{} better predicts stellar abundances, especially for elements C+N and Mn and for stars at super-solar metallicities.
The model makes assumptions, including especially that it fixes some parameters to break degeneracies and improve interpretability; we find that some of the nucleosynthetic implications are dependent upon these detailed parameter choices.
We add a third and fourth process (to make $K=4$), finding that the additional processes give the model more freedom and improve the model's ability to predict the stellar abundances, as expected, but they don't qualitatively change the story. 
The results of \name{} have implications for the formation of the Galaxy disk, the relationship between abundances and ages, and the physics of nucleosynthesis.
\end{abstract}

@@

\begin{deluxetable}{cccc}
    \tablecaption{Fiducial model $\Acc$ and $\AIa$ values for our stellar sample.
    \label{tab:As}}
    \tablehead{
\colhead{APOGEE ID} &	\colhead{[Mg/H]} & \colhead{$\Acc$} &	\colhead{$\AIa$}} 
\startdata
2M00000546+6152107 &	-0.20 &	0.61 &	0.48\\
2M00000866+7122144 &	-0.10 &	0.82 &	0.69\\
2M00001328+5725563 &	0.04 &	1.09 &	1.02\\
2M00001653+5540107 &	0.06 &	1.13 &	0.36\\
2M00001717+6147500 &	-0.23 &	0.60 &	0.39\\
... & ... & ... & ...\\
\enddata
\tablecomments{Full table available online.}
\end{deluxetable}

@@

Such analysis with \name{} parameters will be useful in studying nucleosynthesis, dynamics, disk formation, stellar ages, and much more. However, in this paper we only present fits for a small population with restricted stellar parameters, relative to the full APOGEE sample. 
While \name{} could be fit to the full APOGEE sample, systematic abundance effects with $\teff$ and $\logg$, as well as other abundance artifacts \citep[e.g.,][]{jonsson2020, griffith2021a}, cause the abundance trends to differ across the Hertzsprung-Russell diagram. The best-fit \name{} parameters for the giants would differ from those for the dwarfs. If such systematics could be accounted for (see Sit et al. in prep) we could fit the full APOGEE stellar sample with \name{}, or train \name{} on a subset of high signal-to-noise stars and apply the fits to the full sample. This potential future analysis could reveal additional information about the nucleosynthetic history of our Galaxy and would provide higher signal-to-noise abundance labels for the full sample.
